{RILS}-{ROLS} : Robust Symbolic Regression via Iterated Local Search and Ordinary Least Squares

### 2022-12-01 paper_resources/doc/bmc_jbd/bmc_jbd.tex
Round 2: up to Section 3.3.1

\usepackage{tikz}
%\usepackage{fullpage}
\usepackage{calc}
\usetikzlibrary{positioning,shadows,arrows,trees,shapes,fit}
\usepackage{blindtext}
\usepackage{pgfplots}
%\pgfplotsset{compat=1.16}
\usepackage{pythonhighlight}

\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{url}
\raggedbottom
%\algnewcommand\algorithmicforeach{\textbf{for each}}
%\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}

@@

	\subsection{Ordinary least square method}\label{sec:ols}
Ordinary least square method (OLS) is a linear regression technique. It is based on applying the least-square method to minimize the square residual (error) sum  between actual and predicted values (given by the model). More precisely, given previously introduced dataset $D$ of $n$ points $(\mathbf{x_i}, y_i)$ where each $\mathbf{x_i}$ is $d$-dimensional, the task is to determine linear mapping $\tilde{y} = \mathbf{k} \mathbf{x} + b$, that is coefficients (line slope) $\mathbf{k} = (k_1, \ldots, k_d)$ and $b$ (intercept), so that $ \sum_{i}^{n} (\tilde{y}_i - y_i)^2 $ is minimized. This sum is also know as the sum of squared errors (SSE). There are many methods to minimize SSE. One of the analytical approaches is the calculus-oriented -- it takes into account partial derivatives of SSE w.r.t. $k_j,\; j \in \{1, ..., d\}$:

@@

Now, the algorithm enters the internal loop -- it iterates over the ordered perturbations and aims to find the one which improves the best solution $bs$. But, before comparing candidate perturbation solution $p$ with $bs$, $p$ is first simplified (line~\ref{line:simp}), after which the local search is performed (line~\ref{line:ls}).
Solution simplification is done in a symbolical fashion by popular \texttt{SymPy} Python package~\cite{sympy}.
Local search tries to find local optima expressions close to the given $p$ -- explained in detail in Section~\ref{sec:ls}.  
Finally, the fitness function value of $p$ is compared to the fitness function value of $bs$. If fitness of $p$ is better, $bs$ is updated correspondingly and the internal loop, that goes across ordered perturbations, is immediately terminated (it works in a  \emph{first-improvement} strategy). Otherwise, the next perturbation is probed. 
Note that the probed perturbations are stored in a set denoted by $perturbations_{tried}$. The goal is to avoid checking the same perturbation multiple times (lines \ref{line:avoid1}-\ref{line:avoid2}), i.e. $perturbations_{tried}$ serves as a kind of tabu list, which is known from the Tablu search meta-heuristic, see~\cite{glover1998tabu}.

If some of $s_{perturbations}$ around starting solution $s$ yielded an improvement, $bs$ becomes the starting solution $s$ in the next iteration of the main loop (line~\ref{line:improved}). 
Otherwise, it makes no sense to set starting solution to $bs$, as nothing changed -- the search is being  \emph{trapped} in a local optimum. Randomness is introduced in order to avoid this undesired situation. First, a set of local perturbations around $bs$ is generated ($start_{candidates}$) in the same manner as before (procedure \texttt{All1Perturbations}). If at least one of these was not previously used as a starting solution ($start_{tried}$), a single perturbation from the $start_{candidates} \setminus start_{tried}$ is randomly picked (line~\ref{line:randPick}). There is a minor chance that $start_{candidates} \setminus start_{tried} = \emptyset$. When that happens, the set of starting solution candidates is equal to perturbations of some randomly selected perturbation of $bs$ (lines \ref{line:pert21}-\ref{line:pert22}) -- which effectively means that the perturbations with distance 2 from $bs$ are used. There is a very small chance that these 2-perturbations will have an empty set difference with $start_{tried}$ set of starting solutions. Therefore, 3-perturbations are not considered in our algorithm.

Before returning the final symbolical model, \textsc{RILS}-\textsc{ROLS}  performs the final symbolical simplification and rounding of model coefficients. The latter is sensitive to control parameter $tolerance_{error}$ -- a smaller value means the information loss during rounding is smaller, i.e. a higher number of significant digits is preserved. Note that confidence in the symbolical model is reduced when expected noise in data is higher -- the rule of thumb is: for higher expected noise $tolerance_{error}$ should be higher.

\subsubsection{Fitness function}\label{sec:fitness}

@@

For the (13) competitors, the results are exported from SRBench \url{https://cavalab.org/srbench/results/}, reported in~\cite{la2021contemporary}. 
The maximum computation time allowed for each run of \textsc{RILS}-\textsc{ROLS} is set to 1 hour, while the maximal number of fitness function evaluations is set to 1 million. All other algorithms from SRBench used the time limit that was equal or bellow 8 hours runtime and the same fitness evaluation limit of 1 million (see~\cite{la2021contemporary}). We did not use the full time limit of 8 hours because we empirically concluded that there were almost no improvements after 1 hour \textsc{RILS}-\textsc{ROLS} runtime.

@@

\subsection{Datasets and SRBench}

SRBench (\url{https://cavalab.org/srbench/}) is an open-source benchmarking project which merges a large set of diverse benchmark datasets, contemporary SR methods as well as ML methods around a shared model evaluation and an environment for analysis, see~\cite{la2021contemporary}. SRBench considers two types of symbolic regression problems: 1) \emph{ground-truth} problems, for which the exact model is known, and 2) \emph{black-box} problems, for which the exact model is not known. In our work, we consider only the former one, since \textsc{RILS-ROLS} was not designed to solve \emph{black-box} problems. There are two groups of instances in SRBench \emph{ground-truth} problem set:
\begin{itemize}
	\item \textsc{Feynman} instances are inspired by physics and formulas/models that describes various natural laws.  
	There are 116 instances, where each one consists of $10^5$  samples (see \cite{udrescu2020ai} for more details). Some exact models (equations) of \textsc{Feynman} instances are listed in Table~\ref{tab:Feynamn-Eq}.

@@

In order to check the statistical significance of the results w.r.t.\ obtained exact models and their differences between the competitor approaches, we employed the statistical methodology comparing the results of \textsc{RILS}--\textsc{ROLS} to 13 other approaches over all algorithm runs on the problem instances from the both ground-truth benchmark sets, \textsc{Feynman} and \textsc{Strogatz}. Note that an indicator random variable is involved for each run of each algorithm. More precisely, the score that comes into the statistical observation is an indicator (0 or 1) indicating weather or not an algorithm was able to return the exact model at specific run. More precisely, we involved 1300 results into our statistical evaluation per each algorithm. Note that  results of some of the 13 competitors are not known from SRBench \url{https://cavalab.org/srbench/results/} due to   unknown reasons (most likely algorithm is terminated running out of memory or due to   chosen algorithm configuration settings). For these runs, we simply indicate a 0 as their score.

>>>>

\usepackage{tikz}
%\usepackage{fullpage}
\usepackage{calc}
\usetikzlibrary{positioning,shadows,arrows,trees,shapes,fit}
\usepackage{blindtext}
\usepackage{pgfplots}
%\pgfplotsset{compat=1.16}
\usepackage{pythonhighlight}
\usepackage{fixme}
\fxsetup{status=draft} % <====== add this line
\usepackage{mathtools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{url}
\raggedbottom
%\algnewcommand\algorithmicforeach{\textbf{for each}}
%\algdef{S}[FOR]{ForEach}[1]{\algorithmicforeach\ #1\ \algorithmicdo}

@@

	\subsection{Ordinary least square method}\label{sec:ols}
Ordinary least square method (OLS) is a linear regression technique. It is based on applying the least-square method to minimize the square residual (error) sum  between actual and predicted values (given by the model). More precisely, given   dataset $D$ of $n$ points $(\mathbf{x_i}, y_i)$ where each $\mathbf{x_i}$ is $d$-dimensional, the task is to determine linear mapping $\tilde{y} = \mathbf{k} \mathbf{x} + b$, that is coefficients (line slope) $\mathbf{k} = (k_1, \ldots, k_d)$ and $b$ (intercept), so that $ \sum_{i}^{n} (\tilde{y}_i - y_i)^2 $ is minimized. This sum is also know as the sum of squared errors (SSE). There are many methods to minimize SSE. One of the analytical approaches is the calculus-oriented -- it takes into account partial derivatives of SSE w.r.t. $k_j,\; j \in \{1, ..., d\}$:

@@

Now, the algorithm enters the internal loop -- it iterates over the ordered perturbations and aims to find the one which improves the best solution $bs$. But, before comparing candidate perturbation solution $p$ with $bs$, $p$ is first simplified (line~\ref{line:simp}), after which the local search is performed (line~\ref{line:ls}).
Solution simplification is done in a symbolical fashion by popular  Python package \texttt{SymPy}~\cite{sympy}.
Local search tries to find local optima expressions close to the given $p$ -- explained in detail in Section~\ref{sec:ls}.  
Finally, the fitness function value of $p$ is compared to the fitness function value of $bs$. If fitness of $p$ is better, $bs$ is updated correspondingly and the internal loop, that goes across ordered perturbations, is immediately terminated (it works in a  \emph{first-improvement} strategy). Otherwise, the next perturbation is probed. 
Note that the probed perturbations are stored in a set denoted by $perturbations_{tried}$. The goal is to avoid checking the same perturbation multiple times (lines \ref{line:avoid1}-\ref{line:avoid2}), i.e. $perturbations_{tried}$ serves as a kind of tabu list, which is known from the Tablu search meta-heuristic, see~\cite{glover1998tabu}.

If some of $s_{perturbations}$ around starting solution $s$ yielded an improvement, $bs$ becomes the starting solution $s$ in the next iteration of the main loop (line~\ref{line:improved}). 
Otherwise, it makes no sense to set starting solution to $bs$ -- the search is being  \emph{trapped} in a local optimum. Randomness is introduced in order to avoid this undesired situation. First, a set of local perturbations around $bs$ is generated ($start_{candidates}$) in the same manner as before (procedure \texttt{All1Perturbations}). If at least one of these was not previously used as a starting solution ($start_{tried}$), a single perturbation from the $start_{candidates} \setminus start_{tried}$ is randomly picked (line~\ref{line:randPick}). There is a minor chance that $start_{candidates} \setminus start_{tried} = \emptyset$. When that happens, the set of starting solution candidates is equal to perturbations of some randomly selected perturbation of $bs$ (lines \ref{line:pert21}-\ref{line:pert22}) -- which effectively means that the perturbations with distance 2 from $bs$ are used. There is a very small chance that these 2-perturbations will have an empty set difference with $start_{tried}$ set of starting solutions. Therefore, 3-perturbations are not considered in our algorithm.

Before returning the final symbolical model, \textsc{RILS}-\textsc{ROLS}  performs the final symbolical simplification and rounding of model coefficients. The latter is sensitive to control parameter $tolerance_{error}$ -- a smaller value means the information loss during rounding is smaller, i.e. a higher number of significant digits is preserved. Note that confidence in the symbolical model is reduced when expected noise in data is higher -- the rule of thumb is: for higher expected noise, $tolerance_{error}$ should be higher.

\subsubsection{Fitness function}\label{sec:fitness}
\fxnote{Marko: Dosao dovde!}

@@

For the (13) competitors, the results are exported from \texttt{SRBench} \url{https://cavalab.org/srbench/results/}, reported in~\cite{la2021contemporary}. 
The maximum computation time allowed for each run of \textsc{RILS}-\textsc{ROLS} is set to 1 hour, while the maximal number of fitness function evaluations is set to 1 million. All other algorithms from \texttt{SRBench} used the time limit that was equal or bellow 8 hours runtime and the same fitness evaluation limit of 1 million (see~\cite{la2021contemporary}). We did not use the full time limit of 8 hours because we empirically concluded that there were almost no improvements after 1 hour \textsc{RILS}-\textsc{ROLS} runtime.

@@

\subsection{Datasets and \texttt{SRBench}}

\texttt{SRBench} (\url{https://cavalab.org/srbench/}) is an open-source benchmarking project which merges a large set of diverse benchmark datasets, contemporary SR methods as well as ML methods around a shared model evaluation and an environment for analysis, see~\cite{la2021contemporary}. \texttt{SRBench} considers two types of symbolic regression problems: 1) \emph{ground-truth} problems, for which the exact model is known, and 2) \emph{black-box} problems, for which the exact model is not known. In our work, we consider only the former one, since \textsc{RILS-ROLS} was not designed to solve \emph{black-box} problems. There are two groups of instances in \texttt{SRBench} \emph{ground-truth} problem set:
\begin{itemize}
	\item \textsc{Feynman} instances are inspired by physics and formulas/models that describes various natural laws.  
	There are 116 instances, where each one consists of $10^5$  samples (see \cite{udrescu2020ai} for more details). Some exact models (equations) of \textsc{Feynman} instances are listed in Table~\ref{tab:Feynamn-Eq}.

@@

In order to check the statistical significance of the results w.r.t.\ obtained exact models and their differences between the competitor approaches, we employed the statistical methodology comparing the results of \textsc{RILS}--\textsc{ROLS} to 13 other approaches over all algorithm runs on the problem instances from the both ground-truth benchmark sets, \textsc{Feynman} and \textsc{Strogatz}. Note that an indicator random variable is involved for each run of each algorithm. More precisely, the score that comes into the statistical observation is an indicator (0 or 1) indicating weather or not an algorithm was able to return the exact model at specific run. More precisely, we involved 1300 results into our statistical evaluation per each algorithm. Note that  results of some of the 13 competitors are not known from \texttt{SRBench} \url{https://cavalab.org/srbench/results/} due to   unknown reasons (most likely algorithm is terminated running out of memory or due to   chosen algorithm configuration settings). For these runs, we simply indicate a 0 as their score.
